---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# renv2docker

<!-- badges: start -->

[![Lifecycle:
experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![CRAN
status](https://www.r-pkg.org/badges/version/renv2docker)](https://CRAN.R-project.org/package=renv2docker)

<!-- badges: end -->

Project-specific cache of `{renv}` dependencies in Docker workflows

## Motivation

Past me:

1.  I wanted to use `{renv}` in combination with Docker and didn't yet know how
    to mount my local [`{renv}`
    cache](https://rstudio.github.io/renv/articles/renv.html#cache-1), so the
    build process took **much** longer than I was willing to wait ;-)

2.  This also happened at a time before
    [rocker](https://hub.docker.com/u/rocker) switched to Ubuntu as the
    underlying Linux OS (Thanks to [Eric
    Nantz](https://twitter.com/thercast?lang=en) for pointing that out in your
    [podcast](https://github.com/rpodcast/r_dev_projects)!).

    Before that (AFAIU and AFAIR), it could happen that the compilation process
    slightly differed between Ubuntu and Debian(?). So even though you might
    have properly mounted your local [`{renv}`
    cache](https://rstudio.github.io/renv/articles/renv.html#cache-1) so it was
    available at build time, the resulting builds might have been incompatible
    with your local Linux OS.

    To be honest, I'm not sure how much of this still holds/is necessary after
    the switch to Ubuntu, but I nevertheless liked the idea of having an
    `{renv}` cache that is completely project-specific, so I continued building
    this package out.

    [**TODO: Investigate the rocker history, etc. so I don't tell incorrect
    stuff here** ;-)]

So I wanted to have a faster build experience and came up with this two-step
approach:

**STEP 1**

Build an image that handles all of my package dependencies.

It only builds the dependencies once or when it has to (that is whenever my
`renv.lock` is updated) and caches them back to a subdirectory within my local
package directory (`./renv/cache_docker` [TODO: or was it `./renv/cache` --\>
review this]) for future builds.

I called this image the **D**ependency **C**ache **M**anager (**DCM**).

**STEP 2**

Build the actual image

This image get access to the locally persisted `{renv}` cache from the previous
step as well as `./renv/local` (the place where your package's `.tar.gz` file is
built to when you use the `build.R`; see below).

It calls `renv::restore()` which uses the cache and restores your package by
installing it from source.

## Installation

```{r eval=FALSE}
remotes::install_github("rappster/renv2docker")
```

## Usage

```{r example}
library(renv2docker)
```

1.  Create the necessary environment variables for Docker

```{r}
write_env_vars()
```

Technically, they are mostly used as `ARG`s instead of `ENV`s, but at least for
`PACKAGE_PORT` it would make sense to pass it on as e `ENV`. Still learning
about the nitty-gritty details ;-) See [Docker ARG, ENV and .env - a Complete
Guide](https://vsupalov.com/docker-arg-env-variable-guide/) for more details on
this

1.  Generate `./Dockerfile` and `./renv/Dockerfile`

```{r}
dockerfile <- use_template_dockerfile(open = FALSE)

readLines(dockerfile) %>% cat(sep = "\n")
```

```{r}
dockerfile <- use_template_dockerfile_dcm(open = FALSE)

readLines(dockerfile) %>% cat(sep = "\n")
```

1.  Generate `build.R`

This template ensures that your package is built into `./renv/local` and that
all workflow requirements of `renv2docker` are met regarding the caching of
package dependency builds.

```{r}
build_r <- use_template_build(open = FALSE)

readLines(build_r) %>% cat(sep = "\n")
```

1.  Generate `docker_build.sh` and `docker_run.sh`

```{r}
build_build_sh <- use_template_docker_build(open = FALSE)

readLines(build_build_sh) %>% cat(sep = "\n")
```

```{r}
docker_run_sh <- use_template_docker_run(open = FALSE)

readLines(docker_run_sh) %>% cat(sep = "\n")
```

1.  Modify the desired R script to be run inside the Docker container

Find line

    COPY inst/main.R main.R

and line

    CMD Rscript 'main.R'

and adapt it to your setup.

The default setup

-   expects a file `main.R` within your `inst` directory
-   copies that into the Docker image as file `main.R`
-   Runs `main.R` via `Rscript`

1.  Build your Docker container

Open a terminal/shell within your package/project root directory and type

    ./docker_build.sh

1.  Run your Docker container

<!-- -->

    ./docker_run.sh

## Shout outs

This package was inspired and influenced by

-   [Colin Fay: An Introduction to Docker for R
    Users](https://colinfay.me/docker-r-reproducibility/)
-   [ColinFay/dockerfiler](https://github.com/ColinFay/dockerfiler)
-   [ThinkR/devindocker](https://github.com/ThinkR-open/devindocker)
-   [Eric Nantz: R Dev Projects](https://github.com/rpodcast/r_dev_projects)

My goal is to align this package as much as I can with existing workflows and
best practices around R in combination with Docker.

## TODOs

-   Learn much more about Docker ;-)
-   Thoroughly investigate existing tooling around R and docker - especially the
    packages and workflows mentioned in the shout outs - to see where I can
    improve the package by better aligning it with best practices

## DISCLAIMER

This is another one of my "scratch your own itch" type of projects.

I use Linux ([Pop!\_OS](https://pop.system76.com/)) and hence aligned `{renv}`
and Docker workflows to that platform --\> not sure how much of it is applicable
for MacOS or Windows

Still hope it works for other developers out there as well :-)

## Code of Conduct

Please note that the renv2docker project is released with a [Contributor Code of
Conduct](https://contributor-covenant.org/version/2/0/CODE_OF_CONDUCT.html). By
contributing to this project, you agree to abide by its terms.
